Properties of Winning Iterated Prisoner's Dilemma Strategies.

### 2020-04-25 paper/main.tex
rewrites discussion while trying to keep everyones inputs

This highlights why TFT's early tournament
success fails to generalize -- it never attempts to defect against a cooperating
or exploitable opponent (e.g. Alternator).

This manuscript has explored the performance of \numberofstrategies strategies
of the IPD in a large number of computer tournaments.
We analyzed and extracted the salient features of the best performing strategies
across various tournament types, casting the results in terms of R. Axelrod's
original suggested features of good IPD strategies. Moreover, our results
shed light on the historic performance of TFT, ZDs,
and memory-one strategies generally. Strategies need to match their play
to the cooperativeness of the tournament population and do so in a way that
prevents or minimizes exploitation. Overall we see that complex or clever
strategies can be effective, whether trained against a corpus of possible
opponents or purposely designed to mitigate the impact of noise such as
in the strategy DBS. Further, we showed that while the type of exploitation
attempted by ZDs is not typically effective in tournaments, more
sophisticated strategies capable of selectively exploiting weaker opponents
while mutually cooperating with stronger opponents can be highly successful.
This fact was also indicated numerically by the importance of the strategy
feature SSE in the analysis of strategy features.

% TODO Add discussion about what our results imply and how they are explained
% relative to past performances of TfT and ZDs.

@@

Highly noisy or tournaments with short matches favor less cooperative strategies.
These environments mitigate the value of being nice. Uncertainty enables
exploitation, reducing the ability of maintaining or enforcing mutual cooperation,
while triggering grudging strategies to switch from typically cooperating to
typically defecting. Accordingly, we find that in noisy tournaments
the best performing players cooperate a lower rate than the tournament
population on average. Nevertheless we found some strategies designed or trained
for noisy environments were also highly ranked in noise-free tournaments. This
indicates that strategy complexity is not necessarily a liability, rather it
can confer adaptability to a more diverse set of environments.

In Section~\ref{section:top_performances}, the tournaments results were used to
present the top performances. The data set contained results from four different
settings, and these were also studied individually. In standard tournaments
complex strategies trained using reinforcement learning ranked in the top spots.
Some of these strategies ranked again in the top spots in probabilistic
ending tournaments when a \(p_e\) of less 0.1 was considered and in noisy tournaments
when \(p_n\) was less than 0.1. In probabilistic
ending tournaments \(p_e\) was designed to vary between 0 and 1. It was demonstrated
that for values larger than 0.1, as stated in the Folk Theorem, defecting strategies
were winning the tournaments because there was a high likelihood of the game
ending in the next turn. In tournaments with noise the median ranks of the top
15 strategies had the highest values and the \(r\) distributions were bimodal.
The top rank strategies were performing both well and bad, and this indicates
that in noisy tournaments where the noise can vary substantially,
there were no strategies that can guarantee winning across a range of noise.
However, if the probability of noise was constrained at 0.1 then strategies
designed for noisy tournaments indeed performed well.

So what is the best way of playing the IPD? And is there a single dominant
strategy for the IPD?

There was not a single strategy within the collection of the
\numberofstrategies strategies that managed to perform well in all the
tournaments variations it competed in.
Even if on average a strategy ranked highly in a specific environment this did not
guarantee its success over the different tournament types. However, the results
of sections~\ref{section:top_performances}
and~\ref{section:evaluation_of_performance} demonstrated that there are
properties associated with the success of strategies. A few of the properties that
have been identified by this manuscript's analysis contradict the properties of
Axelrod~\cite{Axelrod1981}. Namely, in Section~\ref{section:top_performances} it
was shown that trained strategies dominated several
tournaments across tournament types, hinting that successful IPD strategies are
often clever or more complex than simple strategies like TFT or ZDs. Most of the
successful strategies highlighted
in~Section~\ref{section:top_performances} were strategies that begin with
cooperation.

Furthermore, in Section~\ref{section:top_performances}
and~\ref{section:evaluation_of_performance} it was shown that envious strategies
performed well. Though these were not the most envious strategies in the
tournaments (ZDs were included), these strategies benefited by being a bit
envious. From Section~\ref{section:evaluation_of_performance} it was concluded
that there is a significant importance in adapting to the environment, and more
specifically in this work, to the mean cooperator. This section also
demonstrated that a strategy should reciprocate, as suggested by R. Axelrod, but in
some environments, such as standard and noisy, it should relax its readiness to
do so.

Thus, the five properties successful strategies need to have in a IPD competition
are:

\begin{itemize}
\item be nice, if the environment isn't noisy and the games are long
    \item be somewhat provocable,
    \item be a little envious,
    \item be clever,
    \item adapt to the environment (including the population of strategies).
\end{itemize}

>>>>

This manuscript explored the performance of \numberofstrategies strategies of
the IPD in \numberofalltournaments computer tournaments. The collection of
computer tournaments presented here is the largest and more diverse collection in the
literature. The \numberofstrategies strategies are drawn from known named
strategies in IPD literature and the computer tournaments include tournaments of
four different settings.

So what is the best way of playing the IPD? And is
there a single dominant strategy for the IPD? There was not a single strategy
within the collection of the \numberofstrategies strategies that managed to
perform well in all the tournaments variations it competed in. Even if on
average a strategy ranked highly in a specific environment this did not
guarantee its success over the different tournament types. Nevertheless, in
Sections~\ref{section:top_performances}
and~\ref{section:evaluation_of_performance} we examined the best performing
strategies across various tournament types and analysed their salient features,
and demonstrated that there are properties associated with the success of
strategies, contradicting the originally suggested properties of R.
Axelrod~\cite{Axelrod1981}.

We showed that complex or \textbf{clever} strategies can be effective, whether
trained against a corpus of possible opponents or purposely designed to mitigate
the impact of noise such as in the strategy DBS. Moreover, we found some
strategies designed or trained for noisy environments were also highly ranked in
noise-free tournaments which reinforces the idea that strategies'
complexity/cleverness is not necessarily a liability, rather it can confer
adaptability to a more diverse set of environments.
We also showed that while the type of exploitation attempted by ZDs is
not typically effective in standard tournaments, \textbf{envious} strategies
capable of both exploiting and not their opponents can be highly successful.
Based on the results of~\cite{Harper2017} this could be because they are
selectively exploiting weaker opponents while mutually cooperating with stronger
opponents. Highly noisy or tournaments with short matches also favoured envious
strategies. These environments mitigated the value of being nice. Uncertainty
enables exploitation, reducing the ability of maintaining or enforcing mutual
cooperation, while triggering grudging strategies to switch from typically
cooperating to typically defecting.

The features analysis of the best performing strategies demonstrated that a
strategy should reciprocate, as suggested by R. Axelrod, but it should relax its
readiness to do so and be more \textbf{generous}. For noisy environments this is
inline with the results of~\cite{Bendor1991, Donninger1986, Molander1985,
Hammerstein1984}, however, we also showed that generosity pays off even in
standard settings, and that in fact the only setting a strategy would want to be
too provocable is when the matches are not long. Forgiveness as defined by R.
Axerlod was not explored here. This was mainly because the two round states were
not recorded during the data collection. That is something we would aim to
incorporate in future work.
The features analysis also concluded that
there is a significant importance in \textbf{adapting to the environment}, and
more specifically, to the mean cooperator. Moreover, the manner in which a
strategy achieves a given cooperation rate relative to the tournament population
average is important. In standard tournaments a strategy would aim to be the
mean the cooperator while in noisy tournaments the best performing players
cooperate a lower rate than the tournament population on average.

This could potentially explain the early success of TFT. TFT naturally achieves
a cooperation rate near $C_{\text{mean}}$ by virtue of copying its opponent's
last move while also minimizing instances where it is exploited by an opponent
(cooperating while the opponent defects), at least in non-noisy tournaments. It
could also explain why Tit For \(N\) Tats does not fare well for $N > 1$ -- it
fails to achieve the proper cooperation ratio by tolerating too many defections.

Similarly, our results could suggest an explanation regarding the intuitively
unexpected effectiveness of memory-one strategies historically. Given that among
the important features associated with success are the relative cooperation rate
to the population average and the four memory-one probabilities of cooperating
conditional on the previous round of play, these features can be optimized by a
memory-one strategy such as TFT. Usage of more history becomes valuable when
there are exploitable opponent patterns. This is indicated by the importance of
SSE as a feature, showing that the first-approximation provided by a memory-one
strategy is no longer sufficient.

@@

Overall, the five properties successful strategies need to have in a IPD competition
based on the analysis that has been presented in this manuscript are:

\begin{itemize}
    \item Be ``nice'' in non-noisy environments or when game lengths are longer
    \item Be provocable but also generous when game lengths are longer
    \item Be a little bit envious
    \item It's ok to be clever
    \item Adapt to the environment (including the population of strategies).
\end{itemize}
